---
title: "Hands-on Time Series Analysis with R"
subtitle: "Chapter 4: Working with zoo and xts"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../../style.css
params:
  date: !r Sys.Date()    
---

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(data.table)
library(xts)
library(TSstudio)
```

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following notes/scripts/plots are inspired by chapter 5 of the book [Hands-on Time Series Analysis with R](https://github.com/PacktPublishing/Hands-On-Time-Series-Analysis-with-R) by Rami Krispin

# 5 Decomposition of time series data

## 5.1 The moving average function
Components of a moving average:

1. rolling window -- generic function that slides across the series to extract sequential subsets.

2. average function -- a simple or weighted average, which is applied on each subset of the rolling window.

### 5.1.1 The rolling window structure
The window structures include:

1. one-sided window -- groups each observation of the series with its past $n-1$ observations. The first $n-1$ observations cannot be smoothed.

2. two-sised window -- groups each observation of the series with past $n_{1}$ and future $n_{2}$ observations where $n$ = $n_{1}$ + $n_{2}$ + 1. If $n$ is odd then $n_{1}$ equals  $n_{2}$.  The $n-1/2$ observations on either end of the series cannot be smoothed. 

### 5.1.2 The average method

1.The simple mean where:

$$\bar{Y} = \frac{\sum^n_{i=1}Y_{i}}{n}$$
2. The weighted average:

$$\bar{Y} = \frac{\sum^n_{i=1}w_{i}Y_{i}}{n}$$
### 5.1.3 The MA attributes
The length of the MA window $n$ is known as the "order".
The cost of MA window are the observations that cannot be averaged because of the definition of the window. For one-sided windows the cost is the first $n-1$ observations; for two-sided, the $n-1/2$ observations on either end of the series.

Main applications of MA are:

1. Noise reduction -- reduces the series variation, smoothing the random noise

2. De-seasonalize -- remove the seasonal component

3. Forecasting -- forecast the future of the series by averaging the past observations

### 5.1.4 The simple moving average
