---
title: "Lesson 1: Time Series Basics"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    theme: darkly
    highlight: default
    css: ../style.css
params:
  date: !r Sys.Date()      
---

```{r, message = FALSE, warning = FALSE}
library(here)
library(ggplot2)
library(data.table)
library(modelr)
library(TSstudio)
library(RplotterPkg)
library(RtsaPkg)

current_dir <- here::here("PennState Course")
```

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following notes, scripts, and plots are following the [Lesson 1: Time Series Basics](https://online.stat.psu.edu/stat510/lesson/1).
</div>

## 1.1 Overview of time series characteristics

### 1.1.1 Type of models
There are two basic types of "time domain" models:
 
1. Models that relate the present value of a series to past values -- ARIMA models (Autoregressive Integrated Moving Average).

2. Ordinary regression models that use time indices a x-variables.

### 1.1.2 Earthquake time series example
<div class="task">
  Task: Read and plot earthquake data.
</div>  

1. Read and convert data to data.frame object::
```{r}
apath <- file.path(current_dir,"data/quakes.dat")
quakes_vec <- scan(apath)
quakes_df <- data.frame(
  date = seq(from = as.Date("1920-01-01"), by = "year", length.out = 99),
  data = quakes_vec
)
```

2. Plot data:
```{r, fig.width=12, fig.height=8}
RplotterPkg::create_scatter_plot(
  df = quakes_df,
  aes_x = "date",
  aes_y = "data",
  title = "Time Series Plot of Quake Counts",
  subtitle = "Over 99 consecutive years",
  #x_limits = c(as.Date("0000-01-01"), as.Date("0100-01-01")),
  #x_major_breaks = seq(as.Date("0000-01-01"), as.Date("0100-01-01"), "10 year"),
  x_date_labels = "%Y",
  rot_y_tic_label = TRUE,
  x_title = "Year",
  y_title = "Quakes Count",
  connect = TRUE,
  pts_fill = "red",
  pts_color = "blue"
)
```

* no consistent trend
* no seasonality
* no obvious outliers
* unsure if variance is constant or not

### 1.1.3 Autoregressive model of order 1 
One of the simplest ARIMA type models is to use a linear model to predict the present value using the value from one previous time or order of 1 called AR(1). The order k indicates the number of previous times ($x_{t-1}, x_{t-2},...x_{t-k}$ to use in the model to estimate the current time $x_t$.

<div class="task">Task: Calculate and plot the $x_t$ (i.e. "*quakes*") versus $x_{t-1}, x_{t-2},...x_{t-4}$  for the quakes time series.</div> 

```{r, fig.width=16, fig.height=8}
quakes_lags_ar_1 <- RtsaPkg::graph_lead_lag(
  df = quakes_df,
  time_col = "date",
  value_col = "data",
  title = "Relationship Between Lags of Quakes and Itself",
  rot_y_tic_label = TRUE,
  show_fit = TRUE
)
```

Appears to have a slight positive linear association as lag 1, so AR(1) model might be a useful model. 

<div class="task">Task: Show the statistical summary for the lag 1 linear regression.</div>  
```{r}
quakes_ar1_model <- quakes_lags_ar_1$models$lag_1
summary(quakes_ar1_model)
```

The model is essentially the ordinary simple linear equation.  
The algebraic expression of the AR(1) model is:
$$x_t = \delta + \phi_1x_{t-1} + w_t$$
where the error terms $w_t \overset{iid}{\backsim} N(0,\sigma_w^2)$.

It appears that the slope coefficient is significantly different from 0, so AR(1) is a helpful predictor.  The $R^2$ is relatively weak so the model won't give us great predictions.

### 1.1.4 The AR(1) fit residual analysis

<div class="task">Task: Plot the fitted values of $x(t)$ with corresponding AR(1) fitted residuals.</div> 

1. Show the structure of *quakes_ar1_model* model variable:
```{r}
str(quakes_ar1_model$model)
```

2. Plot "Predictions" vs "Residuals" from *quakes_ar1_model*:
```{r, fig.width=10, fig.height=8}
RplotterPkg::create_scatter_plot(
  df = quakes_lags_ar_1$predictions[[1]],
  aes_x = "Predictions",
  aes_y = "Residuals",
  title = "Residuals of Predicting Quakes(t) from Quakes(t-1)",
  pts_fill = "red",
  rot_y_tic_label = TRUE
) + ggplot2::geom_hline(aes(yintercept = 0), color = "red", linetype = "dashed", lwd = 1)
```

### 1.1.5 Autocorrelation function (ACF)
The autocorrelation function for a series gives the correlations between the series $x_t$ and $x_{t-1}$, and $x_t$ and $x_{t-2}$ and so on for lags 1,2,3,..k.
The autocorrelation is useful for both identifying the order of the AR portion of ARIMA and for checking the lag correlations across the residual series. The correlations for the residual series should not be significant across all the lags.

<div class="task">Task: Plot 18 lags of the ACF of the residuals from the quakes series.</div>  
```{r, fig.width=10, fig.height=6}
acf_quake_residuals <- RtsaPkg::graph_acf(
  df = quakes_lags_ar_1$predictions[[1]],
  time_col = "DateTime",
  value_col = "Residuals",
  max_lag = 18,
  confid_level = 1.96,
  title = "Autocorrelation for Residuals of Quakes AR(1) Model Fit",
  show_obs = FALSE,
  show_pc = FALSE,
  ac_x_limits = c(1,18),
  ac_y_limits = c(-0.4,0.4),
  ac_y_major_breaks = seq(-0.4,0.4,0.1),
  bold_y = 0.0,
  row_height = 4
)
```

Appears that none of the lags for the models AR(1) residuals are significant.  Residuals usually are theoretically assumed to have an ACF that has correlation = 0 for all lags.

## 1.2 Sample ACF and properties of AR(1) model

### 1.2.1 Stationary series
Definition for *weakly stationary* series:

* The mean $E(x_t)$ is the same for all $t$.

* The variance of $x_t$ is the same for all $t$.

* The covariance (and also correlation) between $x_t$ and $x_{t-1} is the same for all $t$.

Definition of Autocorrelation Function (ACF) for a stationary series:
$$Covariance(x_t,x_{t-k})/SD(x_t)SD(x_{t-k}) = Covariance(x_t,x_{t-k})/Variance(x_t)$$
The denominator in the second formula occurs because the sd of a stationary series is the same at all times.
Many stationary series have recognizable ACF patterns.

### 1.2.2 The first order autoregression model
The algebraic expression of the AR(1) model is:
$$x_t = \delta + \phi_1x_{t-1} + w_t$$
Assumptions:

* $w_t \overset{iid}{\backsim} N(0,\sigma_w^2)$ - errors are independently distributed with a normal distribution
        that has mean 0 and constant variance.
        
* properties of the errors $w_t$ are independent of $x$.

* the series $x_1, x_2,...$ is weakly stationary. A requirement for a stationary AR(1) model is that $|\phi|$ < 1.

### 1.2.3 Properties of the AR(1)

* (theoretical)mean of $x_t$:
$$E(x_t) = \mu = \delta/(1 - \phi_1)$$

* variance of $x_t$:
$$Var(x_t) = \sigma_w^2/(1 - \phi_1^2)$$

* correlation between observations h time periods apart is:
$$\rho_h = \phi_1^h$$

<span class="note">
Note: $\phi_1$ is the slope in the AR(1) model and it is also the lag 1 autocorrelation (h = 1).
</span>

### 1.2.4 Pattern of ACF for AR(1) model

<div class="task">Task: Simulate an AR(1) model with lag 1 autocorrelation of 0.6</div>  
1. Create the AR(1) simulated model:
```{r}
ar_1_positive_ts <-  stats::arima.sim(list(ar = c(0.6)), n = 1000)
RtsaPkg::get_series_info(ar_1_positive_ts)
```
2. Convert the simulated "ts" to a data.frame:
```{r}
ar_1_positive_df <-  RtsaPkg::ts_to_df(ar_1_positive_ts, col_name = "AR(1)")
```

3. Plot the ACF:
```{r,fig.width=10, fig.height=6}
ar_1_positive <- RtsaPkg::graph_acf(
  df = ar_1_positive_df,
  time_col = "DateTime",
  value_col = "AR(1)",
  max_lag = 12,
  confid_level = 1.96,
  ac_x_limits = c(1,12),
  ac_y_limits = c(-0.2,0.8),
  ac_y_major_breaks = seq(-0.2,0.8,0.1),
  title = "ACF for AR(1) with Phi = 0.6",
  show_obs = FALSE,
  show_pc = FALSE,
  show_minor_grids = FALSE,
  bold_y = 0.0,
  row_height = 4
)
```

For AR(1) we see a tapering pattern where for positive $\phi_1$, the ACF exponentially decreases to 0 with increasing lags.

<div class="task">Task: Simulate an AR(1) model with lag 1 autocorrelation of -0.7.</div>  
1. Simulate the AR(1):
```{r}
ar_1_negative_ts <-  stats::arima.sim(list(ar = c(-0.7)), n = 1000)
```

2. Create a data.frame for plotting:
```{r}
ar_1_negative_df <-  RtsaPkg::ts_to_df(ar_1_negative_ts, col_name = "AR(1)")
```

3. Plot the AR(1):
```{r,fig.width=10, fig.height=6}


ar_1_negative <- RtsaPkg::graph_acf(
  df = ar_1_negative_df,
  time_col = "DateTime",
  value_col = "AR(1)",
  max_lag = 12,
  confid_level = 1.96,
  ac_x_limits = c(1,12),
  ac_y_limits = c(-0.8,0.6),
  ac_y_major_breaks = seq(-0.8,0.6,0.1),
  title = "ACF for AR(1) with Phi = -0.7",
  show_obs = FALSE,
  show_pc = FALSE,
  show_minor_grids = FALSE,
  bold_y = 0.0,
  row_height = 4
)
```

The AR(1) with a negative $\phi_1$ also decays exponentially but with the algebraic signs for the autocorrelation alternate between positive and negative.

## 1.3 ACF of the quakes series
<div class="task">Task: Compute and plot the ACF of the quakes time series data set</div>  

1. Re-Read in the quakes data set and create a data.frame:
```{r}
apath <- file.path(current_dir,"data/quakes.dat")
quakes_vec <- scan(apath)
quakes_df <- data.frame(
  date = seq(from = as.Date("1920-01-01"), by = "year", length.out = 99),
  data = quakes_vec
)
```
2. Plot the ACF of the quakes series:
```{r,fig.width=10, fig.height=6}
quakes_acf <- RtsaPkg::graph_acf(
  df = quakes_df,
  time_col = "date",
  value_col = "data",
  max_lag = 12,
  confid_level = 1.96,
  ac_x_limits = c(1,12),
  ac_y_limits = c(-0.2,0.6),
  ac_y_major_breaks = seq(-0.2,0.6,0.1),
  title = "ACF for Quakes Time Series",
  show_obs = FALSE,
  show_pc = FALSE,
  show_minor_grids = FALSE,
  bold_y = 0.0,
  row_height = 4
)
```

The values of the lag 1 to lag 4 autocorrelation are shown below. Compare lag 1 (.5417) with what we found above in the regression.

```{r}
quakes_acf$acf_df[1:5,]
```

The expression for correlation $\rho_h = \phi_1^h$ imples that $lag2 \approxeq lag1^2$. $lag1^2 = .2935$ which is somewhat less than 0.419.  Important point--the sample ACF will rarely fit a perfect theoretical pattern.

## 1.4 Daily cardiovascular mortality time series

<div class="task">Task: Plot the daily cardiovascular mortality time series</div>  
1. Read in the data and create a data.frame:
```{r}
apath <- file.path(current_dir,"data/cmort.dat")
cardio_vec <- scan(apath)
cardio_df <- data.frame(
  date = seq(from = as.Date("1970-01-01"), by = "week", length.out = 508),
  data = cardio_vec
)
```
2. Plot cardio_df:
```{r, fig.width=14, fig.height=8}
RplotterPkg::create_scatter_plot(
  df = cardio_df,
  aes_x = "date",
  aes_y = "data",
  title = "Weekly Cardiovascular Mortality Rate",
  subtitle = "Los Angeles County, 1970-1979",
  x_title = "Month-Year",
  y_title = "Mortality",
  x_major_date_breaks = "6 month",
  x_date_labels = "%b%y",
  y_limits = c(65,135),
  y_major_breaks = seq(65,135,10),
  rot_y_tic_label = TRUE,
  connect = TRUE,
  pts_fill = "red",
  pts_color = "blue"
)
```

Slight downward trend, so series may not be stationary. To de-trend, examine first differences $y_t = x_t - x_{t-1}$.

<div class="task">Task: Take first differences of the cardio time series.</div>  
```{r, fig.width=12, fig.height=8}
cardio_diff <-  RtsaPkg::graph_dif(
  df = cardio_df,
  time_col = "date",
  value_col = "data",
  title = "First Differences for the Cardio Time Series",
  subtitle = "Los Angeles County, 1970-1979",
  x_title = "Month-Year",
  y_title = "Mortality",
  x_major_date_breaks = "6 month",
  x_date_labels = "%b%y",
  show_obs = FALSE,
  show_pts = FALSE
)
```

The trend appears to be removed.

<div class="task">Task: Plot the ACF of the cardio differences.</div>  
1. Show the differences data frame:
```{r}
str(cardio_diff$diff_df)
```

2. Plot the autocorrelation of the first differences:
```{r, fig.width=13, fig.height=6}
cardio_diff_acf <- RtsaPkg::graph_acf(
  df = cardio_diff$diff_df,
  time_col = "DateTime",
  value_col = "Value",
  max_lag = 30,
  title = "Autocorrelation of First Differences for the Cardio Time Series",
  subtitle = "Los Angeles County, 1970-1979",
  ac_x_limits = c(1,30),
  ac_x_major_breaks = seq(1,30,1),
  ac_y_limits = c(-0.8,0.4),
  ac_y_major_breaks = seq(-0.8,0.4,0.1),
  show_obs = FALSE,
  show_pc = FALSE,
  show_minor_grids = FALSE,
  confid_level = 1.96,
  bold_y = 0.0,
  col_width = 12,
  row_height = 4
)
```

Show the first five acf values:
```{r}
cardio_diff_acf$acf_df[(1:5),]
```
The ACF pattern looks like a AR(1) with a negative $lag_1$ autocorrelation.  Also the powers of $lag_1$ are close to the values of the other $lag_2^2$, $lag_3^3$... .*

<div class="task">Task: Access the first differences data.frame from **cardio_diff**.</div> 
```{r}
cardio_diff_df <- cardio_diff$diff_df
```
<div class="task">Task: Plot the relationship of the cardio series with its various lagged series</div>  
```{r, fig.width=16, fig.height=8}
cardio_diff_lag <- RtsaPkg::graph_lead_lag(
  df = cardio_diff_df,
  time_col = "DateTime",
  value_col = "Value",
  title = "CardioDiff(t) vs CardioDiff(t-k) for Cardio Differences Series",
  rot_y_tic_label = TRUE,
  show_fit = TRUE
)
```

Appears to be a slight negative relationship for lag 1 implying an AR(1) model with a negative coefficient.
This is in line with the lag 1 ACF from above with a value of -0.5060.

<div class="task">Task: Apply the summary() function to the model for lag 1 from *cardio_diff_lag$models*.</div>  
```{r}
summary(cardio_diff_lag$models$lag_1)
```


The coefficient *lag_1* is significant at -0.506 but the $R^2$ at 0.256 is weak.
