---
title: "Analysis of Time Series-An Introduction"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    css: ../../style.css
params:
  date: !r Sys.Date()      
---

```{r, message=FALSE, warning=FALSE}
library(data.table)
library(here)
library(hms)
library(ggplot2)
library(grid)
library(gtable)
library(tseries)
library(TSstudio)
library(lubridate)
library(RtsaPkg)
library(RplotterPkg)

current_dir <- here::here("Analysis of Time Series-An Introduction")
```

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following notes/scripts/plots are inspired by the book [The Analysis of Time Series: An Introduction](http://www.ams.sunysb.edu/~xing/tsRbook/index.html) by Chris Chatfield and Haipeng Xing
</div>

Note that sections were skipped that did not have any scripting/plotting requirements.

# Some Representative Time Series
## Classic Beveridge wheat price index series
1. Info on the classic Beveridge wheat price index series `tseries::bev`:

```{r}
data(bev, package = "tseries") # set the data

is.ts(bev)
length(bev)
head(bev, 10)
head(stats::time(bev), 10)
head(stats::cycle(bev), 10)
stats::frequency(bev)
stats::deltat(bev)
stats::start(bev)
stats::end(bev)
```
2. Summary of the `tseries::bev` data set:
```{r}
TSstudio::ts_info(bev)
```
3. Convert the `tseries::bev` time series object (class = "ts") to a data.frame/data.table:
```{r}
wheat_index_dt <- RtsaPkg::ts_to_df(bev, col_name = "PriceIdx")
str(wheat_index_dt)
```
4. Plot the Beveridge wheat price index series:
```{r}
RplotterPkg::create_scatter_plot(
    df = wheat_index_dt,
    aes_x = "DateTime",
    aes_y = "PriceIdx",
    title = "Beveridge Wheat Price Index",
    subtitle = "1500 to 1869",
    x_title = "Year",
    y_title = "Price Index",
    rot_y_tic_label = TRUE,
    pts_size = 0.7,
    pts_color = "purple",
    x_limits = as.Date(c("1499-01-01", "1870-01-01")),
    x_major_breaks = seq.Date(from = as.Date("1450-1-1"), to = as.Date("1900-1-1"), by = "50 year"),
    x_date_labels = "%Y",
    y_limits = c(0, 500),
    show_pts = FALSE,
    connect = TRUE
)
```

## Closing prices of S&P 500 index
1. Read the csv file with closing prices from Jan 4, 1995 to Dec 30, 2016:
```{r}
apath <- file.path(current_dir, "data", "sp500_1995-2016.csv")
s_p_500_dt <- data.table::fread(apath)
str(s_p_500_dt)
```
2. Convert `Date` variable from character to date type.
```{r}
s_p_500_dt[, Date_alt := as.Date(Date, format = "%m/%d/%y")]
str(s_p_500_dt)
```
3. Plot the series:
```{r, fig.width=12}
RplotterPkg::create_scatter_plot(
    df = s_p_500_dt,
    aes_x = "Date_alt",
    aes_y = "Return",
    x_limits = as.Date(c("1995-01-01", "2017-01-01")),
    x_major_date_breaks = "2 year",
    x_date_labels = "%b %Y",
    title = "S & P 500 Closing Index",
    subtitle = "Jan 4, 1995 to Dec 30, 2016",
    x_title = "Day",
    y_title = "Daily Return",
    rot_y_tic_label = TRUE,
    pts_size = 0.7,
    pts_color = "purple",
    show_pts = FALSE,
    connect = TRUE
)
```

## Air temperature in Anchorage, Alaska
1. Read in the csv file of monthly average air temperature in Anchorage, Alaska from 2001 to 2016:
```{r, message=FALSE}
apath <- file.path(current_dir, "data", "alaska_temp_1951-2016.csv")
alaska_temps_dt <- data.table::fread(apath)
str(alaska_temps_dt)
```
2. Convert the `Date` variable to a Date variable:
```{r}
alaska_temps_dt[, Date_alt := paste(substring(as.character(Date), 1, 4), substring(as.character(Date), 5, 6), "01", sep = "-")]
alaska_temps_dt[, Date_alt := as.Date(Date_alt)]
str(alaska_temps_dt)
```
3. Plot the Alaska temperatures starting in 2001:
```{r}
RplotterPkg::create_scatter_plot(
  df = alaska_temps_dt,
  aes_x = "Date_alt",
  aes_y = "Celsius",
  title = "Monthly Average Temperature in Anchorage, Alaska",
  subtitle = "Jan 1, 2001 to Dec 31, 2016 (deg C)",
  x_title = "Date",
  y_title = "Average Temperature",
  rot_y_tic_label = TRUE,
  pts_size = 0.7,
  pts_color = "purple",
  x_limits = as.Date(c("2001-01-01", "2016-12-31")),
  x_major_date_breaks = "1 year",
  x_date_labels = "%Y",
  show_pts = FALSE,
  connect = TRUE
)
```

## Domestic sales of Australian wine
1. Read the csv file of Australian wine in successive quarters from March 1985 to June 2014:

```{r, message=FALSE}
apath <- file.path(current_dir, "data", "aus_wine_sales.csv")
wine_sales_dt <- data.table::fread(apath)
str(wine_sales_dt)
```
2. Rename the variables:
```{r}
wine_sales_dt <- wine_sales_dt[, .(DateChr = V1, Sales = V2)]
str(wine_sales_dt)
```
3. Add a `Date` type column:
```{r}
wine_sales_dt[, Date_alt := paste(substring(DateChr, 5, 8), substring(DateChr, 1, 3), "01", sep = "-")]
wine_sales_dt[, Date_alt := as.Date(Date_alt, format = "%Y-%b-%d")]
str(wine_sales_dt)
```
3. Plot the wine sales:
```{r, fig.width=12}
RplotterPkg::create_scatter_plot(
  df = wine_sales_dt,
  aes_x = "Date_alt",
  aes_y = "Sales",
  title = "Domestic Sales of Australian Wine",
  subtitle = "March 1985 to June 2014 by quarter",
  x_title = "Date",
  rot_y_tic_label = TRUE,
  pts_size = 0.7,
  pts_color = "purple",
  x_limits = as.Date(c("1985-01-01", "2015-01-01")),
  x_major_date_breaks = "2 year",
  x_date_labels = "%Y",
  connect = TRUE
)
```

## Total population and birth rate for the US
1. Read the csv file of total population and birth rate for the US from 1965 to 2015:
```{r, message=FALSE}
apath <- file.path(current_dir, "data", "US_pop_birthrate.csv")
birth_pop_dt <- data.table::fread(apath)
str(birth_pop_dt)
```
2. Rename variables:
```{r}
birth_pop_dt <-  birth_pop_dt[, .(DateChr = DATE, Population = populaton, BirthRate = `birthrate(per1000)`)]
str(birth_pop_dt)
```
3. Add a `Date` type:
```{r}
birth_pop_dt[,`:=`(Year_alt = lubridate::parse_date_time2(DateChr, "m/d/y", cutoff_2000 = 16L), Population = Population/10000000)]
str(birth_pop_dt)
```
4. Put the series into a "long" form where *Population* and *BirthRate* are separate observations:
```{r}
birth_pop_dt <- data.table::melt(birth_pop_dt, measure.vars = c("Population", "BirthRate"), variable.name = "Measure", value.name = "Value")
str(birth_pop_dt)
```
4. Create separate plots for  *BirthRate* and *Population* across *Year_alt*:
```{r,fig.height=8, fig.width=12}
plot_birth_pop <- RplotterPkg::multi_scatter_plot(
  df = birth_pop_dt,
  factor_var = "Measure",
  factor_x = "Year_alt",
  columns = 1,
  aes_y = "Value",
  title = "Total Yearly BirthRate/Population for the US",
  subtitle = "(Pop x 10 million) 1960 to 2015",
  x_title = "Year",
  y_titles = c("Rate", "Pop"),
  rot_y_tic_label = TRUE,
  pts_color = "violet",
  pts_fill = "blue",
  pts_size = 0.7,
  x_major_breaks = seq(from = as.POSIXct("1960-01-01"), to = as.POSIXct("2016-12-31"), by = "4 year"),
  x_date_labels = "%Y",
  show_pts = FALSE,
  connect = TRUE,
  row_height = 3,
  col_width = 11
)
```

## Prices and volumes of IBM stocks
1. Read the transaction prices and volumes of IBM stocks:
```{r, message=FALSE}
apath <- file.path(current_dir, "data", "taq_trade_ibm_100104.txt")
ibm_trade_dt <- data.table::fread(apath)
str(ibm_trade_dt)
```
2. Rename columns:
```{r}
ibm_trade_dt <- ibm_trade_dt[, .(Time = V1, Price = V2, G127 = V3, Corr = V4, Cond = V5, Ex = V6, Size = V7)]
str(ibm_trade_dt)
```
3. Convert the time:
```{r}
ibm_trade_dt[, `:=`(Time_alt = hms::as_hms(Time), Size = Size/10)]
str(ibm_trade_dt)
```
4. Filter the time for 3 minutes between 09:35:00 to 09:38:00:
```{r}
ibm_trade_3min_dt <- ibm_trade_dt[Time_alt > as_hms("09:35:00") & Time_alt < as_hms("09:38:00")]
str(ibm_trade_3min_dt)
```
5. Separate plots of *Price* and *Size*:
```{r, fig.width=12, fig.height=10}
price_plot <- RplotterPkg::create_stick_plot(
  df = ibm_trade_3min_dt,
  base_val = 131.4,
  aes_x = "Time_alt",
  aes_y = "Price",
  title = "IBM Prices",
  x_title = "Time",
  y_title = "Stock Price",
  y_limits = c(131.4, 132),
  y_major_breaks = seq(131.4,132,0.2),
  show_minor_grids = FALSE,
  x_date_labels = "%H-%M-%S"
)
volume_plot <- RplotterPkg::create_stick_plot(
  df = ibm_trade_3min_dt,
  aes_x = "Time_alt",
  aes_y = "Size",
  title = "IBM Volume",
  x_title = "Time",
  y_title = "Stock Volume",
  x_date_labels = "%H-%M-%S"
)

layout <- list(
  plots = list(price_plot, volume_plot),
  rows = c(1,2),
  cols = c(1,1)
)

stock_plot <- RplotterPkg::multi_panel_grid(
  layout = layout,
  col_widths = 10,
  row_heights = c(4,4),
  title = "IBM Transaction Prices and Volume (x10)",
  subtitle = "09:35:00 - 09:38:00",
)
```
