---
title: "Analysis of Time Series-An Introduction"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    css: ../../style.css
params:
  date: !r Sys.Date()    
---

```{r, message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(grid)
library(gtable)
library(RtsaPkg)
library(RplotterPkg)
```

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following notes/scripts/plots are inspired by the book [The Analysis of Time Series: An Introduction](http://www.ams.sunysb.edu/~xing/tsRbook/index.html) by Chris Chatfield and Haipeng Xing
</div>

Note that sections were skipped that did not have any scripting/plotting requirements.

# 3 Some Linear Time Series Models
## 3.4 Purely Random Processes

<div class="task">Task: Compute and display a purely random series and its autocorrelation.</div>
1. Create 500 points of a random series with 0 mean and variance of 1.

```{r}
random_df <- data.frame(
    Time = seq(1,500,1),
    Series = rnorm(n = 500, mean = 0, sd = 1)
)
```

2. Plot the random series along with its autocorrelation:
```{r, fig.width=10, fig.height=11}
acf_random <- RtsaPkg::graph_acf(
    df = random_df,
    time_col = "Time",
    value_col = "Series",
    max_lag = 30,
    title = "Autocorrelation from a Random Normal Distribution",
    x_title = "Index",
    y_title = "Value",
    confid_level = 1.96,
    row_height = 3.5
)
```

## 3.5 Random Walks
Suppose that $Z_t$ is a purely random process with mean $\mu$ and variance $\sigma_Z^2$. A process $X_t$ is said to be a random walk if
$$X_t = X_{t-1} + Z_t$$
The process starts at zero when $t = 0$ so that 
$$X_1 = Z_1$$
and 
$$X_t = \sum_{i=1}^tZ_i$$
$X_t$ is the sum of a random process and we find that $E(X_t) = t\mu$ and $Var(X_t) = t\sigma_z^2$ since the $Z_i$'s are independent. As the mean and variance change with $t$, the process is non-stationary.

It is interesting to note that the first differences of a random walk, given by:
$$\Delta X_t = X_t - X_{t-1} = Z_t$$
forms a purely random process, which is therefore stationary.
The best-known examples of time series which behave like random walks are share prices on successive days. A model which often gives a good approximation to such data is:

<div class="note">*share price on day* $t$ = *share price on day* $t-1$ + random error</div>

<div class="task">Task: Create and plot a random walk process.</div>  
1. Create a data.frame with random walk:

```{r}
random_walk_df <- data.frame(
    Time = seq(1,500,1),
    Series = cumsum(rnorm(n = 500, mean = 0, sd = 1))
)
```
2. Plot the random walk series along with its autocorrelation:
```{r, fig.width=10,fig.height=11}
acf_random_walk <- RtsaPkg::graph_acf(
    df = random_walk_df,
    time_col = "Time",
    value_col = "Series",
    max_lag = 30,
    title = "Autocorrelation from a Random Walk Normal Distribution",
    x_title = "Index",
    y_title = "Value",
    confid_level = 1.96,
    row_height = 3.5
)
```

<div class="note">The random walk appears to be non-stationary with the mean and variance changing with $t$ and that there is a significant correlation at lag 1.</div>


<div class="task">Task: Take first differences of the above random walk and plot the ACF to check if the significant lag 1 still holds up.</div>  

1. Plot first differences of the random walk:
```{r,fig.width=10, fig.height=6}
random_walk_diff <- RtsaPkg::graph_dif(
  df = random_walk_df,
  time_col = "Time",
  value_col = "Series",
  title = "First Differences of a Random Walk",
  show_pts = FALSE,
  show_obs = FALSE
)
```

2. Plot the ACF/PACF of the random walk's first order differences:
```{r, fig.width=10, fig.height=10}
random_walk_diff_plot <- graph_acf(
  df = random_walk_diff$diff_df,
  time_col = "DateTime",
  value_col = "Value",
  max_lag = 30,
  title = "ACF & PACF of Random Walk First Differences",
  confid_level = 1.96,
  show_obs = FALSE,
  bold_y = 0.0,
  row_height = 4
)
```
<div class="note">Appears that we are back to a stationary random process without any significant lags.</div>

## 3.6 Moving Average Processes

### 3.6.1 Stationarity and autocorrelation function of an MA process


