---
title: "Analysis of Time Series-An Introduction"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    css: ../../style.css
params:
  date: !r Sys.Date()      
---

```{r, message=FALSE, warning=FALSE}
library(data.table)
library(here)
library(ggplot2)
library(grid)
library(gtable)
library(tseries)
library(RtsaPkg)
library(RplotterPkg)

current_dir <- here::here("Analysis of Time Series-An Introduction")
```

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following notes/scripts/plots are inspired by the book [The Analysis of Time Series: An Introduction](http://www.ams.sunysb.edu/~xing/tsRbook/index.html) by Chris Chatfield and Haipeng Xing
</div>

# 2 Basic Descriptive Techniques

## 2.5 Series that contain a trend and no seasonal variation

### 2.5.2 Filtering
<div class="task">
  Task: Obtain the Beveridge wheat price data set and plot it. Note that the data set `bev` is a time series object of class "ts".
</div>

```{r}
data(bev, package = "tseries") # set the data
str(bev)
```
<div class="task">Task: Plot the `bev` data set. </div>
1. Convert the `bev` time series object (class = "ts") to a data.frame using `RtsaPkg::ts_to_df()`:
```{r}
wheat_index_dt <- RtsaPkg::ts_to_df(bev, col_name = "PriceIdx")
str(wheat_index_dt)
```
2. Call `RplotterPkg::create_scatter_plot()` to define a scatter plot of `bev`:
```{r, fig.width=10, fig.height=10}
wheat_index_plot <- RplotterPkg::create_scatter_plot(
    df = wheat_index_dt,
    aes_x = "DateTime",
    aes_y = "PriceIdx",
    title = "Beveridge Wheat Prices",
    x_title = "Year",
    y_title = "Index",
    connect = TRUE
)
```
3. Call `RtsaPkg::graph_ma()`to define a simple moving average plot:
```{r, fig.width=10, fig.height=6}
wheat_index_sma <- RtsaPkg::graph_ma(
  wheat_index_dt, 
  time_col = "DateTime",
  value_col = "PriceIdx", 
  title = "Beveridge Wheat Prices",
  subtitle = "Simple Moving Average (window = 7)",
  window_n = 7,
  x_axis_title = "Year",
  y_axis_title = "Index",
  show_pts = FALSE,
  palette_colors = c("black","red"))
```
4. Call `RtsaPkg::graph_ma()`to define a weighted moving average plot:
```{r, fig.width=10, fig.height=6 }
wheat_index_weighted <- RtsaPkg::graph_ma(
  df = wheat_index_dt, 
  time_col = "DateTime",
  value_col = "PriceIdx",
  ma_type = "wma",
  window_n = 7,
  title = "Beveridge Wheat Prices",
  subtitle = "Weighted Moving Average (window = 7)",
  x_axis_title = "Year",
  y_axis_title = "Index",
  show_pts = FALSE,
  palette_colors = c("black","red")
)
```
5. Combine the plots *wheat_index_plot*, *wheat_index_sma*, *wheat_index_weighted* into a single plot:
```{r, fig.width=10,fig.height=12}
layout <- list(
  plots = list(wheat_index_plot, wheat_index_sma$plots, wheat_index_weighted$plots),
  rows = c(1,2,3),
  cols = c(1,1,1)
)
multi_plot <- RplotterPkg::multi_panel_grid(
  layout = layout,
  col_widths = 10,
  row_heights = c(3.5,3.5,3.5),
  title = "Beveridge Wheat Prices",
  subtitle = "Moving Average Smoothing 1500 to 1869"
)
```

## 2.6 Series that contain a trend and seasonal variation

<div class="task">
  Task: Decompose the quarterly domestic sales of Australian wine time series into seasonal, trend, and irregular components by calling `RtsaPkg::graph_decompose()`.</div>  
1. Read in the data:
```{r, message=FALSE}
apath <- file.path(current_dir, "data", "aus_wine_sales.csv")
wine_sales_dt <- data.table::fread(apath)
str(wine_sales_dt)
```
2. Rename the variables:
```{r}
wine_sales_dt <- wine_sales_dt[, .(DateChr = V1, Sales = V2)]
str(wine_sales_dt)
```
3. Add a `Date` type column:
```{r}
wine_sales_dt[, Date_alt := paste(substring(DateChr, 5, 8), substring(DateChr, 1, 3), "01", sep = "-")]
wine_sales_dt[, Date_alt := as.Date(Date_alt, format = "%Y-%b-%d")]
str(wine_sales_dt)
```
4. Convert `wine_sales_dt` to a time series object:
```{r}
wine_sales_ts <- ts(data = wine_sales_dt$Sales, frequency = 4, start = c(1985, 2))
str(wine_sales_ts)
```
5. Call `RtsaPkg::graph_decompose()` and return both the original, seasonal, trend, random values and their associated plots:
```{r, fig.width=12,fig.height=11}
wine_sales_decomposed <- RtsaPkg::graph_decompose(
    series_ts = wine_sales_ts,
    title = "Decomposed Time Series of Domestic Sales of Australian Wine",
    subtitle = "June 1985 to June 2014 by quarter",
    x_title = "Year",
    y_title = "Sales",
    x_major_date_breaks = "2 year",
    x_date_labels = "%Y"
)
```

## 2.7 Autocorrelation and the Correlogram
### 2.7.2 Interpreting the correlogram
<div class="task">
  Task: Create a random series and plot both the series and its correlogram using `RtsaPkg::graph_acf()`.
</div>  

1. Create a series data frame from a random normal distribution and plot the autocorrelation (ACF) and partial autocorrelation (PACF):
```{r}
set.seed(1)
random_df <- data.frame (
  time = seq(1:400),
  value = rnorm(400)
)
```
2. Create the plot:
```{r,fig.width=10,fig.height=11}
acf_random <- RtsaPkg::graph_acf(
  df = random_df,
  time_col = "time",
  value_col = "value",
  title = "Random Series and Autocorrelations",
  x_title = "Index",
  y_title = "Random Value",
  confid_level = 1.96,
  row_height = 3.5
)
```
<div class="note">The acf's hover around zero and don't pass the 95% confidence level.</div>

<div class="task">Task: Plot the autocorrelation from a series that has a trend and is non-stationary.</div>

1. Create the non-stationary series:
```{r}
set.seed(1)
# set up the plot dataframe
nonstat_df <- data.frame(
  time = seq(1:400),
  value = cumsum(rnorm(400))
)
```

2. Plot the trending non-stationary series:
```{r,fig.width=10,fig.height=11}
acf_nonstat <- RtsaPkg::graph_acf(
  df = nonstat_df,
  time_col = "time",
  value_col = "value",
  title = "Non-stationary Random Time Series",
  x_title = "Index",
  y_title = "Non-stat Value",
  confid_level = 1.96,
  row_height = 3.5
)
```
<div class="note">Appears that the lag at 1 is significant.</div>

<div class="task">Task: Plot the autocorrelation of monthly observations on air temperaature in Anchorage, Alaska.</div>

1. Read the csv file:
```{r, message=FALSE}
apath <- file.path(current_dir, "data", "alaska_temp_1951-2016.csv")
alaska_temps_dt <- data.table::fread(apath)
str(alaska_temps_dt)
```
2. Convert the `Date` variable to a Date variable:
```{r}
alaska_temps_dt[, Date_alt := paste(substring(as.character(Date), 1, 4), substring(as.character(Date), 5, 6), "01", sep = "-")]
alaska_temps_dt[, Date_alt := as.Date(Date_alt)]
str(alaska_temps_dt)
```
3. Plot the autocorrelation:
```{r, fig.width=10, fig.height=11}
acf_alaska <- RtsaPkg::graph_acf(
  df = alaska_temps_dt,
  time_col = "Date_alt",
  value_col = "Celsius",
  title = "Autocorrelation for Monthly Alaska Air Tempertures",
  x_title = "Year",
  y_title = "Temperture degC",
  obs_x_major_date_breaks = "10 year",
  obs_x_date_labels = "%Y",
  confid_level = 1.96,
  row_height = 3.5
)
```
