---
title: "Detect Changes in Timeseries Data"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../style.css
params:
  date: !r Sys.Date()    
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
With some random data with three distinct group means in the timeseries, we will compare R package methods
`changepoint::cpt.mean()` and `bcp::bcp()` for locating the groups.

The following R scripts and graphs were inspired from the article [Detect Changes in Timeseries Data](https://www.r-bloggers.com/2021/03/detect-the-changes-in-timeseries-data/?utm_source=feedburner&utm_medium=email&utm_campaign=Feed%3A+RBloggers+%28R+bloggers%29)
</div>  

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(data.table)
library(ggplot2)
library(RplotterPkg)
library(changepoint)
library(bcp)
```

## Setup and plot random data
1. Setup data:
```{r}
set.seed(5)
sleep_v <- rnorm(100, 60, 5)
awake_v <- rnorm(200, 75, 8)
exercise_v <- rnorm(50, 135, 12)
my_heart_rate <- c(sleep_v, awake_v, exercise_v)

data_dt <- data.table(
    time = 1:length(my_heart_rate),
    heart_rate = my_heart_rate
)
```
2. Plot data:
```{r, fig.width = 12, fig.height = 12}
RplotterPkg::create_scatter_plot(
  df = data_dt,
  aes_x = "time",
  aes_y = "heart_rate",
  title = "Heartrate Data",
  subtitle = "Three distinct periods",
  connect = T,
  line_size = 2.0,
  line_color = "blue",
  axis_text_size = 13,
  show_pts = F
)
```

## Detect changes with `changepoint::cpt.mean()`
1. Call the method:
```{r}
change_points <- changepoint::cpt.mean(data = my_heart_rate, penalty = "None", method = "SegNeigh", Q = 3)
pts_dt <- data.table(
  x = c(0, 101, 300, 350),
  y = c(60.15, 75.14, 132.7, 132.7)
)
```
2. Plot the means for the groups identified:
```{r, fig.width = 12, fig.height = 12}
RplotterPkg::create_scatter_plot(
  df = data_dt,
  aes_x = "time",
  aes_y = "heart_rate",
  title = "Heartrate Data",
  subtitle = "Three distinct periods",
  line_size = 2.0,
  line_color = "blue",
  axis_text_size = 13
) + geom_segment(x = pts_dt$x[[1]], y = pts_dt$y[[1]], xend = pts_dt$x[[2]], yend = pts_dt$y[[1]], color = "red") +
    geom_segment(x = pts_dt$x[[2]], y = pts_dt$y[[2]], xend = pts_dt$x[[3]], yend = pts_dt$y[[2]], color = "red") +
    geom_segment(x = pts_dt$x[[3]], y = pts_dt$y[[3]], xend = pts_dt$x[[4]], yend = pts_dt$y[[3]], color = "red")
```

## Detect changes with `bcp::bcp()`
1. Call the method:
```{r}
bcp_lst <- bcp::bcp(y = my_heart_rate)
bcp_pts_dt <- data.table(
  time = 1:length(bcp_lst$posterior.mean),
  heart_rate = as.numeric(bcp_lst$posterior.mean)
)
```

2. Plot the means for the groups identified:
```{r, fig.width = 12, fig.height = 12}

RplotterPkg::create_scatter_plot(
  df = data_dt,
  aes_x = "time",
  aes_y = "heart_rate",
  title = "Heartrate Data",
  subtitle = "Three distinct periods",
  line_size = 2.0,
  line_color = "blue",
  axis_text_size = 13
) + geom_line(data = bcp_pts_dt, color = "red")
```
