---
title: "Hands-on Time Series Analysis with R"
subtitle: "Chapter 2: Working with Date and Time Objects"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    css: ../../style.css
params:
  date: !r Sys.Date()  
---

```{r, message=FALSE, warning=FALSE}
library(here)
library(data.table)
library(lubridate)

current_dir <- here::here("Hands-on Time Series Analysis with R")
```

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following notes/scripts/plots are inspired by chapter 2 of the book [Hands-on Time Series Analysis with R](https://github.com/PacktPublishing/Hands-On-Time-Series-Analysis-with-R) by Rami Krispin

Also at [Amazon](https://www.amazon.com/Hands-Time-Analysis-forecasting-visualization-ebook/dp/B07B41P2HZ/ref=sr_1_2?dchild=1&keywords=Hands-On+Time+Series+Analysis+with+R&qid=1592992293&s=books&sr=1-2)
</div>

# 2 Working with Date and Time Objects
## 2.1 Date and time objects in R

Create date and time objects
```{r}
(date <- Sys.Date())
(time_ct <- Sys.time())
(time_lt <- as.POSIXlt(time_ct))
```
Show their class.
```{r}
class(date)
class(time_ct)
class(time_lt)
```
Unclass to show their internal storage.
```{r}
time_un_ct <- unclass(time_ct) # returns numeric distance from origin point
time_un_lt <- unclass(time_lt) # returns a list of length 11 with $year $mon $mday $hour, $min, $sec, etc
time_un_lt$zone
time_un_lt$year
```
## 2.2 Creating date and time objects
From string to Date object. The following are in the form of the ISO 8601 standard:
```{r}
(date <- as.Date("2014-5-12"))
(time_ct <- as.POSIXct("2014-5-12 20:05:35", tz = "EST"))
```
## 2.3 Importing date and time objects
The origin point varies with software. R uses Jan 1, 1970 and uses the ISO8601 standard.

## 2.4 Reformatting and converting date objects
<div class="task">Task: Read the `dates_format.csv` file and convert the dates. </div>
1. Read the data:
```{r, warning = FALSE, message = FALSE}
apath <- file.path(current_dir, "data", "dates_formats.csv")
dates_dt <- data.table::fread(apath)
str(dates_dt)
```
2. Convert the time columns:
```{r}
dates_new_dt <- dates_dt[, .(Japanese = as.Date(Japanese_format),
                             US_format = as.Date(US_format, format = "%m/%d/%Y"),
                             US_long_format = as.Date(US_long_format, format = "%A, %B %d, %Y"),
                             CA_mix_format = as.Date(CA_mix_format, format = "%B %d, %Y"),
                             SA_mix_format = as.Date(SA_mix_format, format = "%d %B %Y"),
                             NZ_format = as.Date(NZ_format, format = "%d/%m/%Y"),
                             Excel_Numeric_Format = as.Date(Excel_Numeric_Format, origin = as.Date("1899-12-30")))]
str(dates_new_dt)
```
## 2.5 Reformatting and conversion of time objects
Using POSIX classes which are an extension of Date.
```{r}
time_str <- "2018-12-31 23:59:59"
time_posix_ct <- as.POSIXct(time_str) # gives the number of seconds since Jan 1, 1970
class(time_posix_ct)
```
If the format of the input object does not follow the YYY-m-d H:M:s structure you will have to use the format argument.
```{r}
time_US_str <- "Monday, December 31, 2018 11:59:59 PM"
(time_posix_us_ct <- as.POSIXct(time_US_str, format = "%A, %B %d, %Y %I:%M:%S %p"))
```
## 2.6 Creating a date or time index
Using  `base::seq.Date()` or  `base::seq.POSIXt()` to create a sequence of time index.
```{r}
daily_index <- seq.Date(from = as.Date("2016-01-01"), to = as.Date("2018-12-31"), by = "day")
head(daily_index)
```
Using the length argument--a POSIXct example with a length of 48 hours.
```{r}
hourly_seq <- seq.POSIXt(from = as.POSIXct("2018-06-01"), by = "hours", length.out = 48)
str(hourly_seq)
```
## 2.7 Manipulation of date and time with the lubridate package
Converting a complex time object not in the ISO 8601 standard.
We had to use the "format = " argument to convert this string:
```{r}
time_US_str <- "Monday, December 31, 2018 11:59:59 PM"
```
Using the `lubridate::ymd_hms()` conversion function:
```{r}
(time_lubridate <- lubridate::mdy_hms(time_US_str, tz = "EST")) # note that hours is 24 hour
```

Note that  `lubridate::mdy_hms()` is following the input string's pattern.  Lubridate offers 12 different patterns of letters for conversion of the input string.
In the `dates_formats.csv` file we can use lubridate's patterns `ymd()`, `mdy()`, `dmy()`:
```{r}
apath <- file.path(current_dir, "data", "dates_formats.csv")
dates_dt <- data.table::fread(apath)
head(dates_dt)
```
Convert the times using methods from `lubridate::`.
```{r}
dates_new_dt <- dates_dt[, .(
    Japanese = ymd((Japanese_format)),
    US_format = mdy(US_format),
    US_long_format = mdy(US_long_format),
    CA_mix_format = mdy(CA_mix_format),
    SA_mix_format = dmy(SA_mix_format),
    NZ_format = dmy(NZ_format),
    Excel_Numeric = lubridate::as_date(Excel_Numeric_Format, origin = as.Date("1899-12-30"))
)]
str(dates_new_dt)
```
Note that for the Excel_Numeric_Format we used as_data() or as_datetime() from lubridate with origin = as.Date("1899-12-30").

## 2.8 Utility functions for date and time objects
Extracting specific time components using lubridate.
```{r}
time_obj <- lubridate::mdy_hms("Monday, December 31, 2018 11:59:59 PM")
# extract the day of the year:
(lubridate::yday(time_obj))
# extract the day of the quarter
(lubridate::qday(time_obj))
# extract the day of the month
(lubridate::day(time_obj))
# extract the day of the week as an ordered factor character string
(lubridate::wday(time_obj, label = TRUE))
```
You can also modify the time object:
```{r}
time_obj <- ymd("2012-05-24")
(day(time_obj) <- 23)
time_obj
```
