---
title: "Hands-on Time Series Analysis with R"
subtitle: "Chapter 4: Working with zoo and xts Objects"
output: 
   html_document:
    toc: yes
    toc_depth: 4
    css: ../../style.css
params:
  date: !r Sys.Date()  
---
```{r, message=FALSE, warning=FALSE}
library(knitr)
library(xts)
library(zoo)
library(data.table)
library(TSstudio)
library(RplotterPkg)
library(RtsaPkg)
library(RplotterPkg)
```

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following notes/scripts/plots are inspired by chapter 4 of the book [Hands-on Time Series Analysis with R](https://github.com/PacktPublishing/Hands-On-Time-Series-Analysis-with-R) by Rami Krispin

# 4 Working with zoo and xts Objects
## 4.1 The *zoo* class

"zoo" is comprised of two elements:

1. Data structure: A vector or matrix format
2. Index vector: Stores the series observation's corresponding index

An example of "zoo" is the Brent Crude Oil Prices series from TSstudio package:
```{r}
RtsaPkg::get_series_info(TSstudio::EURO_Brent)
```
### 4.1.1 The "zoo" class attributes
"zoo" can handle both regular and irregular spaced time series data.
Use `class(zoo time series object)` in determining regular/irregular:
```{r}
class(TSstudio::EURO_Brent)
```
If the "zoo" time series object is regular we can apply `frequency()` and `cycle()`:
```{r}
stats::frequency(TSstudio::EURO_Brent)
head(cycle(TSstudio::EURO_Brent), 12)
stats::start(TSstudio::EURO_Brent)
stats::end(TSstudio::EURO_Brent)
head(stats::time(TSstudio::EURO_Brent), 12)
```
### 4.1.2 The index of the "zoo" object
Unlike the index of the "ts" class, the "zoo" index supports the use of external date and time objects as the series index.

<div class="task">Task: Use the `zoo::index()` function to view the first indexes of TSstudio::EURO_Brent</div>

```{r}
head(zoo::index(TSstudio::EURO_Brent))
```
The "zoo" index is an independent object with its own `attributes()` and `class()` functions:
```{r}
class(zoo::index(TSstudio::EURO_Brent))
attributes(zoo::index(TSstudio::EURO_Brent))
```
You can convert the "zoo" index to other classes as long as it is applicable:
```{r}
EURO_Brent_date_zoo <- TSstudio::EURO_Brent  # make a copy
zoo::index(EURO_Brent_date_zoo) <- as.Date(zoo::index(TSstudio::EURO_Brent))
head(EURO_Brent_date_zoo)
class(index(EURO_Brent_date_zoo))
```
### 4.1.3 Working with date and time objects
"zoo" package contains a number of utility functions for converting time.  For example `as.yearmon()` and `as.yearqtr()` can convert "Date" objects to "yearmon" or "yearqtr" objects:
```{r}
monthly_dates <- seq(as.Date("2019-01-01"), length.out = 12, by = "month")
head(monthly_dates)
monthly_yearmon <- zoo::as.yearmon(monthly_dates)
head(monthly_yearmon)
```
Note that if the "Date" object has a day component then it is lost with the above conversion.

### 4.1.4 Creating a "zoo" object
Creating a "zoo" object follows the same logic as creating a "ts" object:

1. Convert an existing time series object using the `as.zoo()` function.
2. Create a "zoo" object with the `zoo()` function using a numeric vector as input

<div class="task">Task: Convert a "data.frame" object to a "zoo" object with the `zoo()` function. </div>

1. Using the *US_indicators* data set from TSstudio::
```{r}
data(US_indicators)
str(US_indicators)
```
2. Set the frequency argument of `zoo()` to 12:
```{r}
vehicle_sales_zoo <- zoo(x = US_indicators$`Vehicle Sales`, frequency = 12)
class(vehicle_sales_zoo)
stats::frequency(vehicle_sales_zoo)
head(vehicle_sales_zoo)
head(zoo::index(vehicle_sales_zoo),10)
class(zoo::index(vehicle_sales_zoo))
zoo::is.regular(vehicle_sales_zoo, strict = TRUE)
```
Note that the index is a sequence of ordered integers of class "yearmon" and is a regular time series.  A strictly regular time series is any series where all the observations of the series are equally spaced.

Both "zoo" and "xts" time series can have time index classes of "Date", "POSIXct", "chron", "yearmon", "yearqtr", and "DateTime" classes.

The *order_by* argument of `zoo()` can be used to set the *index* operator of the output object. 
<div class="task">Task: Set the *Vehicle Sales* data set to a "zoo" object with the *order.by* argument set to its "Date" variable.</div>

```{r}
vehicle_sales_zoo <- zoo(x = US_indicators$`Vehicle Sales`, order.by = US_indicators$Date, frequency = 12)
head(vehicle_sales_zoo)
class(index(vehicle_sales_zoo))
head(index(vehicle_sales_zoo))
zoo::is.regular(vehicle_sales_zoo, strict = TRUE)
```
Note that *order_by* preserves the class of "Date" as *Date*. Using the "Date" as an index creates some ambiguity, as the distance between each observation could be 28, 29, 30, or 31 days. The series is strictly irregular.

### 4.1.5 Working with multiple time series objects
<div class="task">Task: From the data.frame *US_indicators*, create a "zoo" object with both "Vehicle Sales" and "Unemployment Rate" variables.</div>

```{r}
US_indicators_zoo <- zoo(
  x = US_indicators[,c("Vehicle Sales", "Unemployment Rate")],
  frequency = 12,
  order.by = US_indicators$Date
)
RtsaPkg::get_series_info(US_indicators_zoo)

class(index(US_indicators_zoo))
zoo::is.regular(US_indicators_zoo, strict = FALSE)
```
## 4.2 The "xts" class
"xts" is an extension of the "zoo" class with additional attributes.  Any of the "zoo" functions can be applied to the "xts" object.

<div class="task">Task: Review the "xts" TSstudio::Michigan_CS data set</div>
```{r}
data(TSstudio::Michigan_CS)
RtsaPkg::get_series_info(TSstudio::Michigan_CS)
class(TSstudio::Michigan_CS)
```
Note that the class of the series is both "xts" and "zoo" and carries both class functions.
<div class="task">Task: Use some "zoo" functions on the "xts" object.</div>
1. Using `class()` and `frequency()`:
```{r}
class(index(TSstudio::Michigan_CS))
frequency(TSstudio::Michigan_CS)
```
2. Using `head()` which outputs a matrix:
```{r}
head(TSstudio::Michigan_CS)
```
3. Using `is.regular()`:
```{r}
zoo::is.regular(TSstudio::Michigan_CS)
```
One of the improvements of the "xts" class is the matrix format structure of the object as opposed to the vector of the "zoo" class.
<div class="task">Task: Show the matrix format of the univariate`TSstudio::Michigan_CS` "xts" object.</div>
```{r}
head(TSstudio::Michigan_CS)
```
<div class="task">Task: Create an "xts" multivariate time series object from the US_indicators data.frame using `xts()`.</div>
```{r}
US_indicators_xts <- xts(x = US_indicators[ ,c("Vehicle Sales", "Unemployment Rate")],
  frequency = 12,
  order.by = US_indicators$Date
)
head(US_indicators_xts)
```
### 4.2.1 The "xts" class attributes
For the univariate case, "zoo" has a vector structure; "xts" has a matrix structure. The uniqueness of "xts" is in its functionality.

### 4.2.2 The xts functionality
#### 4.2.2.1 The periodicity function
<div class="task">Task: Retrieve the periodicity of the TSstudio::Michigan_CS "xts" object.</div>
```{r}
xts::periodicity(TSstudio::Michigan_CS)
```
#### 4.2.2.2 Manipulating the object index
<div class="note">Note: `indexClass()` is deprecated. Use `tclass()`.</div>
`xts::tclass()` both sets and retrieves the index class of an "xts" object.
<div class="task">Task: Retrieve the index class of TSstudio::Michigan_CS "xts" object.</div>
```{r}
xts::tclass(TSstudio::Michigan_CS)
```
<div class="task">Task: Convert the index of an "xts" object from *yearmon* to *Date*.</div>
<div class="note">Note: `convertIndex()` returns  a modified "xts" object and does not alter the original.</div>
```{r}
xts::tclass(TSstudio::Michigan_CS)
Michigan_CS_date_xts <- xts::convertIndex(TSstudio::Michigan_CS, "Date")
xts::tclass(Michigan_CS_date_xts)
```
<div class="task">Task: Retrieve the month component of the index.</div> 
```{r}
head(xts::.indexmon(TSstudio::Michigan_CS), 12)
```
<div class="task">Task: Change the format of the index.</div> 
<div class="note">Note: `indexFormat()` is deprecated. Use `tformat()`.</div>
```{r}
Michigan_CS_mdy_xts <- TSstudio::Michigan_CS
xts::tformat(Michigan_CS_mdy_xts) <- "%m-%d-%Y"
head(Michigan_CS_mdy_xts,13)
xts::tclass(Michigan_CS_mdy_xts)
```
#### 4.2.2.3 Subsetting an xts object based on the index properties
<div class="task">Task: Extract from an "xts" object a subset of data based on its index.</div>
Using the multivariate *US_indicators_xts* object created above get the first 12 months:
```{r}
vehicle_sales_xts <- US_indicators_xts$`Vehicle Sales`[1:12]
TSstudio::ts_info(vehicle_sales_xts)
```


## 4.3 Manipulating the zoo and xts objects

### 4.3.1 Merging time series objects
In merging the first step is to review the indexes of two merging objects. As an example Michigan_CS (an "xts" object) with TSstudio::EURO_Brent (a "zoo" object):
```{r}
class(zoo::index(TSstudio::EURO_Brent))
xts::tclass(TSstudio::Michigan_CS)
```

Second step is to review any overlap:
```{r}
RtsaPkg::get_series_info(TSstudio::Michigan_CS)
RtsaPkg::get_series_info(TSstudio::EURO_Brent)
```
There is overlap, so the third step is to use `merge.xts()` function:
```{r}
merge_outer_xts <- xts::merge.xts(Michigan_CS_xts = TSstudio::Michigan_CS, EURO_Brent_zoo = TSstudio::EURO_Brent, join = "outer")
TSstudio::ts_info(merge_outer_xts)
head(merge_outer_xts["1987"])
```

### 4.3.2 Rolling windows
<div class="task">Task: Use `zoo::rollapply()` to calculate a 3 month, moving average for the TSstudio::EURO_Brent *zoo* series.</div>
Review TSstudio::EURO_Brent:
```{r}
RtsaPkg::get_series_info(TSstudio::EURO_Brent)
```

```{r}
EURO_Brent_3ma <- zoo::rollapply(TSstudio::EURO_Brent, width = 3, FUN = mean)
RtsaPkg::get_series_info(EURO_Brent_3ma)
```
<div class="note">Note: The new series is missing the first and last observations from the original series. We cannot create a window for the first and last observations.</div>

### 4.3.3 Creating lags
<div class="task">Task: Use the `stats::lag()` to create a 3 month lag of the TSstudio::EURO_Brent "zoo" object.</div>
```{r}
EURO_Brent_lag3 <- stats::lag(TSstudio::EURO_Brent, k = -3)
RtsaPkg::get_series_info(EURO_Brent_lag3)
```
<div class="note">Note: We have lost the first 3 months of the original series.</div>
<div class="task">Task: Merge both the lag series with the original series.</div>
```{r}
EURO_Brent_lag_3_zoo <- zoo::merge.zoo(TSstudio::EURO_Brent, EURO_Brent_lag3)
head(EURO_Brent_lag_3_zoo)
```
### 4.3.4 Aggregating the zoo and xts objects
<div class="task">Task: Use `stats::aggregate()` to aggregate the TSstudio::EURO_Brent "zoo" object from monthly to quarterly.</div>
```{r}
RtsaPkg::get_series_info(EURO_Brent)
EURO_Brent_qtr_zoo <- stats::aggregate(TSstudio::EURO_Brent, by = zoo::as.yearqtr, FUN = sum)
RtsaPkg::get_series_info(EURO_Brent_qtr_zoo)
head(EURO_Brent_qtr_zoo)
```
<div class="task">Task: Aggregate TSstudio::EURO_Brent using `lubridate::year()`.</div>
```{r}
EURO_Brent_yr_zoo <- stats::aggregate(TSstudio::EURO_Brent, by = lubridate::year, FUN = sum)
head(EURO_Brent_yr_zoo)
```

### 4.3.5 Plotting zoo and xts objects
If we are using ggplot2 then the "zoo" and "xts" objects must be converted to a "data.frame".
The key function to use is `zoo::fortify.zoo()`.
<div class="task">Task: Plot with ggplot2 the "zoo" TSstudio::EURO_Brent.</div>
1. Convert the "zoo" object:
```{r}
EURO_Brent_df <- zoo::fortify.zoo(TSstudio::EURO_Brent)
head(EURO_Brent_df)
```
2. Rename columns:
```{r}
colnames(EURO_Brent_df) <- c("Index", "EURO_Brent")
```
3. Plot the "data.frame":
```{r}
RplotterPkg::create_scatter_plot(
  df = EURO_Brent_df,
  aes_x = "Index",
  aes_y = "EURO_Brent",
  x_major_date_breaks = "1 month",
  x_date_labels = "%Y-%m"
)
```

## 4.4 xts or ts -- which one to use
<blockquote>..."xts" objects has more benefits compared to "ts" objects, since most of the forecasting models in R support only "ts" objects. </blockquote>

